Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-03 drafts/hybrid_zone.tex
added small note about deme size to Assumptions section
peter edited 2/3 through

\title{Blocks of extended ancestry around selected loci in hybrid zones}
\author{}
\date{} % delete this line to display the current date

@@

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity to better understand the mechanisms underlying reproductive isolation and the process of speciation. \yb{Some form of selection against hybrid genotypes is likely common in many hybrid zones. Here, we use a combination of analytical theory and explicit forward simulations} to gain a better understanding of how this impacts patterns of introgression \yb{across genomic and geographic} space. \alisa{We describe the properties of clines across the genomes as they are formed, as well as correlations in ancestry. Clines are expected to form over time $1/s$, where $s$ is the strength of selection against hybrids. Linked clines persist over a  physical scale of $1/T$, where $T$ is the age of the hybrid zone. Additionally, we find that locally disadvantageous alleles move at speed $\sigma\sqrt{s}$ (where $\sigma$ is the variance in dispersal), and tend to exist as relatively recent, small families. The length of continuous tracts of ancestry provide an additional source of information; within a population, blocks of ancestry surrounding single-locus incompatibilities} are longer than the genome-wide average block length, especially when removed from the center of the hybrid zone.  The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

}

@@

%\plr{
%Main points, make sure they appear in the abstract:
%clines establish over time $1/s$, extend along genome to at least $1/T$;
%locally disadvantageous alleles exist as recent, small families (moving at speed $\sigma/\sqrt{s}$);
%haplotype chunks longest around introgressing selected bits, flanked by shorter chunks.
%}

@@

The process of speciation commonly involves allopatric divergence in relative isolation \citep{Coyne2004}.  
When formerly isolated populations come back into contact 
before reproductive isolation is complete, gene flow is still possible. 
As two such populations interbreed, tracts of
individuals' genomes can be labeled according to which ancestral population they are inherited from,
and local interbreeding and migration of such populations
create a gradient of ancestry assignments across geographic space,
centered on their point of contact \citep{Barton1985}. 
%\plr{(check that neutral clines discussed in barton '85)}
If the populations are sufficiently diverged,
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
leading to substantial correlations in ancestry between distant markers.
The influence of selection on linked loci also prevents the mixing of ancestry at regions surrounding these loci. %the effect of which decreases with physical genomic distance \citep{Barton1986,Barton1983}. 
The mixing of ancestries of a given genome is mediated by 
the process of recombination, forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, Baird2003}.   
Patterns of correlated ancestry in admixed populations has been previously described under neutral conditions both in single populations and in contact zones \citep[e.g.][]{Gravel2012,Hellenthal2014,Sedghifar2015}.

While for much of the genome, gene flow will eventually erase 
both geographic clines in ancestry, and strong correlations in ancestry across linked markers, 
subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. 
% Variation in \yb{ the action of } selection against migrant alleles within the genome results in genome-wide heterogeneity in cline width, as the frequency of foreign ancestry surrounding targets of selection is reduced \citep{Barton1979}. 
% While these patterns of co-ancestry have been described in multi-locus systems, the \yb{expected length of ancestral tracts around selected loci in admixing populations remains poorly understood}\alisa{, especially in the geographic context of hybrid zones.}
\yb{While empirical patterns of the block length distribution in hybrid zones have rarely been used to inform the history of selection,}  \alisa{the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection.

The extensive history of population genetic modeling of hybrid zones has focused largely on the equilibrium states of ancestry clines \citep{Barton1979a,Barton1986}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
This provides some expectation for how selection shapes covariance in ancestry at a single locus over \emph{geographic space}.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry over \emph{genomic space}.

Covariance in ancestry across a genome provides information about the number of surviving recombination events that have taken place between ancestral genotypes and therefore provides information about the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. Our goal is to understand how selection against hybrids interacts with dispersal and recombination to produce observed patterns of ancestry around selected sites in recently established hybrid zones. In addition to interpreting non-equilibrium patterns of diversity in hybrid zones, this could be especially important for understanding the impact of selection on the extent of mixing of ancestral genomes and, consequently, the maintenance of species distinctness. In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
If, on the other hand, hybrids have non-zero fitness, some amount of introgression can occur, and providing information about the extent of hybridization and timing of secondary contact \citep{Price2009, Hellenthal2014}. 
\yb{In this latter case, heterogeneous patterns of ancestry across the genome potentially allows researchers to identify putative targets of selection in hybridizing populations.}  
\yb{To date, researchers have done so by examining heterogeneous clines in ancestry across loci \citep{Porter1997, Gompert2012}, but have not made use of the information contained in the length of ancestry blocks around putatively selected loci.}

Under a model of selection against hybrids, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression in surrounding genomic regions. We therefore expect \yb{a deficit of} short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus is expected to be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to be a recent migrant from the other side of the hybrid zone, and therefore will not have experienced many recombination events.

Using a combination of theory and simulated hybrid zones, we present a model of genome-wide patterns coancestry as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. For this, we build on the neutral framework presented in \citet{Sedghifar2015} to derive the expected distribution of ancestry block lengths across a contact zone.  
\yb{The theoretical expectations developed in this manuscript can aid in the identification of loci under selection in hybridizing populations.}

%\plr{explain relationship to \citet{Sedghifar2015}}

\plr{cite selection-equals-subdivided-population stuff}

@@

As in \citet{Sedghifar2015}, 
we gain insight by considering where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to to the $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same;
recombination events may cause us to follow more than one lineage back to determine the block's ancestry,
and the block will be of ancestry $A$ if all lineages trace back to the $A$ side.
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
As the identity of the selected allele determines how selection acts on the haplotype,
and linked variation can only move between the $A$ and the $B$ background in heterozygotes,
a key factor in these models is the density and fecundity of heterozygotes.

\begin{figure}
\includegraphics{figs/BM_schematic}
\caption{A cartoon representation of a haplotype under the model. The black line represents the path, backwards in time, of a site under selection ($L_B$) that is sampled in the present day. The center of the hybrid zone is defined at geographic position $x=0$, such that ancestry A segregates at lower frequency for $x>0$. Here, because of negative selection, $L_B$ migrates rapidly back to $x>0$ from a position $x<0$. Red dots represent recombination events along the entire haplotype. When a recombination separates two chunks of the sampled haplotype takes place, a branching event occurs and these splitting lineages are represented by colored lines (here the initial haplotype is broken up into 3 chunks over 100 generations). Subsequent recombination events can switch the selected background to which a chunk is linked.  Here, yellow lineages are linked to $L_A$ , and blue lineages are linked to $L_B$. Based on position 100 generations after contact, from right to left the chunks are of ancestry $(A,B,B)$.}\label{Fig:schematic}
\end{figure}

%As an approximation, we consider a cline of narrow width (look up cline width for underdominance, Barton/May?), such that the frequency of migrant alleles is exponential and negligible at distance away from the center of the hybrid zone. This means that recombination events that occur away from the center of the zone do not result in a change in $a$.

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citet{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on numerous approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. 
%\alisa{I don't think I understand the independence from $\sigma$?}
%\plr{It gets set up at speed $\sigma$ over a distance proportional to $\sigma$.}

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is known.
We seek to first understand how lineages of sampled individuals behave at the selected locus,
from which lineages at linked loci require only a minor modification.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $\sigma^2 \grad \log p(x,t)$ (i.e., it moves at speed $\sigma^2$ in the potential $-\log p$, note that we use ``drift'' as directional movement as in the terminology of diffusion, rather than its population genetic meaning).  % YB for clarity
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x+r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$.

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction. 
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

Note that this description gives more information than the steady-state solution,
since in that analysis only the ratio $\sigma/\sqrt{s}$ appears.
Here we see that lineages with $\sigma=10$ and $s=.16$ move quite differently
to lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form.

\paragraph{Lineages at linked loci}
The behavior of a lineage at a linked locus is similar to a selected locus,
except that when in heterozygotes the lineage may move between backgrounds:
if we follow back through time a lineage at a locus we find linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left),
but if it spends time in the clinal region where recombination may happen,
it may be found to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.

Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ (same as previously described movement of a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (for $B$ to $A$).

We can describe this process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B$ the first time $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \begin{cases}
             \sigma^2 \grad \log(p(X_\tau,\tau)) d\tau + \sigma dB_\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \grad \log(1-p(X_\tau,\tau)) d\tau + \sigma dB_\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        % \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A \} &= \epsilon r (1-p(X_\tau,\tau)) + O(\epsilon^2) \\
        % \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B \} &= \epsilon r p(X_\tau,\tau) + O(\epsilon^2)  .
        \P\{ T_B &> u \given Z_\tau = A \} = \exp\left( - r \int_0^u (1-p(X_s,s)) ds \right) \\
        \P\{ T_A &> u \given Z_\tau = B \} = \exp\left( - r \int_0^u p(X_s,s) ds \right) .
    \end{aligned}
\end{align}

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to compute expected clines in ancestry,
i.e., the expected proportion of individuals who inherit ancestry from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In other words,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$G^A = \sigma^2 \grad \log(p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$
and
$G^B = \sigma^2 \grad \log(1-p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$,
and then equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \dot q_A &= G^A q_A + r (1-p) (q_A-q_B) \\
    \dot q_B &= G^B q_B + r (1-p) (q_B-q_A) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of $A$ and $B$ lineages, respectively.

>>>>

% other title ideas:
% {Blocks of extended ancestry around selected loci in hybrid zones}
\title{Beyond clines: lineages and haplotype blocks in hybrid zones}
\author{}
\date{} % delete this line to display the current date

@@

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to better understand the mechanisms underlying reproductive isolation and the process of speciation. 
Here, we use a combination of analytical theory and explicit forward simulations
to gain a better understanding of how selection against hybrid gentypes impacts patterns of introgression 
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  physical scale of $1/T$, where $T$ is the age of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side they originated on at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location. 
The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}
}

@@

@@

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
As two such populations interbreed, each portion of each individual's genome 
can be labeled according to which ancestral population is is inherited from,
and local interbreeding and migration of such populations
creates a gradient of ancestry assignments across geographic space,
centered on their point of contact \citep[reviewed in][]{Barton1985}. 
%\plr{(check that neutral clines discussed in barton '85)}
If the populations are sufficiently diverged,
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, Baird2003},
leading to substantial correlations in ancestry (and hence, genotype) between distant markers.
Patterns of correlated ancestry in admixed populations has been previously described under neutral conditions both in single populations and in contact zones \citep[e.g.][]{Gravel2012,Hellenthal2014,Sedghifar2015}.

While for much of the genome, gene flow will eventually erase 
both geographic clines in ancestry and strong correlations in ancestry across linked markers, 
subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. 
Studies of hybrid zones have primarily been based on single-site statistics such as clines,
but today, 
the availability of dense genotype data in combination with linkage information 
makes it now possible to use patterns of the block length distribution
to better understand the processes operating in hybrid zones.

\plr{Literature: Haldane, Nagylaki, Slatkin, \citep{May1975,Slatkin1973}}
\plr{Better summary of what Barton did.}
Population genetic modeling of hybrid zones has focused largely on the 
equilibrium states of clines at selected loci \citep{Barton1979a,Barton1986}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
This provides some expectation for how selection shapes covariance in ancestry at a single locus over geographic space.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry across the genome.

The arrangement of segments of genome inheriting from particular ancestral populations
provides information about the number of recombination events that have taken place between ancestral genotypes, 
and therefore provides information about the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. 
In addition to interpreting non-equilibrium patterns of diversity in hybrid zones, 
this could be especially important for understanding the impact of selection on the mixing of ancestral genomes and, consequently, the maintenance of species distinctness. 
In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
If hybrids have non-zero fitness, some amount of introgression can occur, providing information about the extent of hybridization and timing of secondary contact \citep{Price2009, Hellenthal2014}. 
The resulting heterogeneous patterns of ancestry across the genome have been used to identify putative targets of selection in hybridizing populations,
by examining heterogeneous clines in ancestry across loci \citep{Porter1997, Gompert2012}.

Under a model of selection against hybrids, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression in surrounding genomic regions. We therefore expect a deficit of short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus is expected to be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to have been recently inherited from the other side of the hybrid zone, and therefore will not have
as many ancestors of the locally common type as do neutral haplotypes.

Using a combination of theory and simulated hybrid zones, we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. 
In \citet{Sedghifar2015} we described how lineages move across a neutral zone of secondary contact,
obtaining predictions for distributions of lengths of blocks of ancestry, which we fit to several datasets.
In this work, we extend the same framework to the case of selection against heterozygotes at a single locus,
by describing how lineages move conditional on the frequencies of the selected allele
and allowing lineages to recombine between selected background 
(analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).

@@

As in \citet{Sedghifar2015}, 
we gain insight by considering where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to a spatial position $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same;
recombination events may cause us to follow more than one lineage back to determine the block's ancestry,
and the block will be of ancestry $A$ if all lineages trace back to the $A$ side.
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
As the identity of the selected allele determines how selection acts on the haplotype,
and linked variation can only move between the $A$ and the $B$ background in heterozygotes,
a key factor in these models is the density and fecundity of heterozygotes.

%%%%%%%%
\begin{figure}
\includegraphics{figs/BM_schematic}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    i.e., showing where the ancestors lived from whom the haplotype was inherited (horizontal axis),
    and how long ago they lived (vertical axis).
    The segment contains the $B$ alelle at the selected locus.
    The black line shows the lineage at the selected site in the sampled individual in the present day (at $\tau=0$).
    The center of the hybrid zone is at geographic position $x=0$; in the region $x>0$, ancestry $A$ is less common,
    and because of negative selection, the selected lineage spends little time there.
    Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
    these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
    or can switch lineages at linked sites between selected backgrounds 
    (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
    Here the initial haplotype is broken up into 3 chunks over 100 generations,
    which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations after contact, 
    from left to right the segments are of ancestry $(A,B,B)$.
}\label{Fig:schematic}
\end{figure}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citet{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, despite being slowed down somewhat by selection against heterozygotes.

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele, $p(x,t)$, is known.
We seek to first understand how lineages of sampled individuals behave at the selected locus,
from which lineages at linked loci require only a minor modification.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

Note that this description gives more information than the steady-state cline $p(x)$,
which depends only on $\sigma/\sqrt{s}$.
Here we see that lineages with $\sigma=10$ and $s=.16$ move quite differently
to lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form.

\paragraph{Lineages at linked loci}
The behavior of a lineage at a linked locus is similar to a selected locus,
except that when in heterozygotes the lineage may move between backgrounds:
if we follow back through time a lineage at a locus we find linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left),
but if it spends time in the hybrid zone, where recombination between selected backgrounds may happen,
it may be found to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.

Suppose we sample an allele today at a locus at recombination distance $r$ from the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(just as previously described for a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (from $B$ to $A$).

We can describe this process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B$ the first time $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \begin{cases}
             \sigma^2 \grad \log(p(X_\tau,\tau)) d\tau + \sigma dB_\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \grad \log(1-p(X_\tau,\tau)) d\tau + \sigma dB_\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        % \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A \} &= \epsilon r (1-p(X_\tau,\tau)) + O(\epsilon^2) \\
        % \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B \} &= \epsilon r p(X_\tau,\tau) + O(\epsilon^2)  .
        \P\{ T_B &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        \P\{ T_A &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
    \end{aligned}
\end{align}

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit ancestry from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$G^A = \sigma^2 \grad \log(p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$
and
$G^B = \sigma^2 \grad \log(1-p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$,
so that equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \dot q_A &= G^A q_A + r (1-p) (q_B-q_A) \\
    \dot q_B &= G^B q_B + r p (q_A-q_B) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of lineages of selected alleles of ancestry $A$ and $B$, respectively.
